---
title: "Final Project Summary"
author: "Lyn Brush"
date: "2026-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Research Topic: Spotify Song Statistics\
Team Members: Lyn (Kalyna) Brush\
\
Data Description:\
> From website: This is a dataset of Spotify tracks over a range of 125 different genres. Each track has some audio features associated with it. The data is in CSV format which is tabular and can be loaded quickly.\
> There doesn't seem to be any cleaning that needs to be done, although less specific genre groups could be created as it's own column (rock, dance, etc). There are some E values that could be difficult to work with.\

Link: https://www.kaggle.com/datasets/maharshipandya/-spotify-tracks-dataset?resource=download \
\

Project idea / research questions:\
> I'm planning to investigate the averages for columns like "acousticness" and "speechiness" by genre, as well as popular keys, bpms, and the most common duration. I also want to see how these things affect the popularity score.\
> I will likely use a lot of histograms and box plots to see averages, and scatterplots to see trendlines.
```{r}

```
