---
title: "Analysis of Spotify Statistics"
author: "Lyn Brush, Huu Phuoc Tran"
date: "2026-05-12"
output: github_document
---

## Introduction

The goal of this project is to explore the Spotify Tracks dataset. Analysis of this topic could potentially be useful to musical artists, labels, and music theorists to determine trends across genres and to find what consumers might want.

We will explore the following research questions:

1.  How do the descriptive columns (time_signature, tempo, etc) affect the popularity of a track?
2.  What is the average of all of the other descriptive variables for each genre?
3.  How do mood-related Spotify audio features differ across selected music genres?

## Data

This dataset is from Kaggle, linked here: (<https://www.kaggle.com/datasets/maharshipandya/-spotify-tracks-dataset>). It contains data on 114 thousand songs over 125 different genres. The data is directly downloadable in the csv format. Here is each column and its purpose:

#### Identification Columns

-   track_id: An identification code given by Spotify.
-   artists: The names of the artists who performed the track. If there are multiple, they are separated by a ";".
-   album_name: The name of the album the track is on.
-   track_name: The name of the track.

While these columns are helpful for finding specific tracks, we didn't need to use them for our purposes.

#### Descriptive Columns

-   popularity: The popularity of a track between 1-100. This is calculated with an algorithm based on the total number of plays and recency of those plays.

-   duration_ms: The track length in milliseconds.

-   explicit: Boolean value determining whether it has explicit lyrics or not (True = has explicit lyrics).

-   key: The key of the track. Numerical values represent different keys in ascending order (0 = C, 1= C#/Db, etc)

-   loudness: The loudness of a track in decibels.

-   mode: Indicates whether the track is in a major (1) or minor (0) key.

-   tempo: The tempo (speed) of the track in beats per minute.

-   time_signature: The time signature (beats per measure) of the track. The given number is over 4 (so a value of 3 would translate to 3/4).

-   track_genre: The genre of the track.

#### Mood Columns

These columns are on a scale of 0.0 to 1.0, with 1.0 representing a song that represents the column the most. For example, a 0.9 score in danceability would mean the song is very danceable.

-   danceability: How suitable a track is for dancing based on elements such as tempo and rhythm.

-   energy: How energetic a track is based on intensity, speed, and loudness.

-   speechiness: How much spoken word is in the track.

-   acousticness: How acoustic (non-electronic) the track is.

-   instrumentalness: How little vocal content is included.

-   liveness: How confident the data is that this track was recorded live.

-   valence: How positive-sounding the track is.

Thankfully, the data was already sufficiently cleaned, so we didn't need to do any cleaning on our part.

## Results

#### Loading the Dataset
```{r}

library(readr)
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
dataset <- read_csv('~/Final/data/dataset.csv')

```

#### How do the descriptive columns (time_signature, tempo, etc) affect the popularity of a track?

##### Correlation between Popularity and Duration:

```{r}
cor(dataset$popularity, dataset$duration_ms, use = "complete.obs")

model <- lm(popularity ~ duration_ms, data = dataset)

plot(dataset$duration_ms, dataset$popularity,
     xlab = "Duration (ms)",
     ylab = "Popularity",
     main = "Popularity vs Duration")
abline(model, col = "red")
```

The correlation between the two (-0.007) is basically negligible at such a low number. However, there is a very slight decrease in popularity as songs get longer. This makes sense with the logic of streaming, as longer songs will take longer to stream, and therefore gain streams slower than shorter songs. The graphic supports this, and shows that most songs are clustered around the 3 minute mark, the traditional song length for most genres.

##### Correlation between Popularity and Explicitness

```{r}
cor(dataset$popularity, ifelse(dataset$explicit, 1, 0), use = "complete.obs")

tapply(dataset$popularity, dataset$explicit, mean, na.rm = TRUE)


boxplot(popularity ~ explicit, data = dataset,
        xlab = "Explicit",
        ylab = "Popularity",
        col = c("lightpink", "lavender"),
        main = "Popularity by Explicit Content")

t.test(popularity ~ explicit, data = dataset)
```
With a correlation coefficient of 0.044, explicitness is slightly more relevant to popularity, although it is still not a good indicator. The average popularity score for a non-explicit song is 32.93, and for an explicit song, it's 36.45. The above boxplot shows the discrepancy between the two; if a song is explicit, it is slightly more likely to have a higher popularity.

##### Correlation between Popularity and Key

```{r}

dataset$key <- as.factor(dataset$key)

tapply(dataset$popularity, dataset$key, mean, na.rm = TRUE)

boxplot(popularity ~ key, data = dataset,
        xlab = "Musical Key",
        ylab = "Popularity",
        col = c("lightpink", "plum", "thistle", "lavender", "lightsteelblue2","lightskyblue1"),
        main = "Popularity by Key")

```

While key is displayed using numbers, it's actually a categorical variable. Because of this, a traditional correlation coefficient can't be found. Instead, we found the average popularity of each of the 12 keys to determine the most and least popular keys. The most popular is 4, which corresponds to the key of E, with an average popularity score of 34.64; this is followed closely by 2 (D), with a score of 34.37. The least popular key is 10, which corresponds to A#/Bb, and has an average popularity score of 32.46.

##### Correlation between Popularity and Time Signature

```{r}
dataset$time_signature <- as.factor(dataset$time_signature)

tapply(dataset$popularity, dataset$time_signature, mean, na.rm = TRUE)

boxplot(popularity ~ time_signature, data = dataset,
        xlab = "Time Signature",
        ylab = "Popularity",
        col = c("lightpink", "plum", "thistle", "lavender", "lightsteelblue2"),
        main = "Popularity by Time Signature")
```

Time signature must be treated like a categorical variable, just like key. According to the boxplot, the most popular time signature is 0/4, with a popularity score of 36.97. It also has the most outliers. This is because 0/4 isn't really a time signature, and these tracks aren't songs. One track with 0/4 time signature, "Weißes Rauschen: Schlafmittel", is a white noise track. The most popular actual time signature is 4/4, with a score of 33.67. This makes sense, since 4/4 is by far the most used time signature across most genres.

##### Correlation between Popularity and Tempo

```{r}

cor(dataset$popularity[dataset$tempo != 0],
    dataset$tempo[dataset$tempo != 0],
    use = "complete.obs")


plot(dataset$tempo[dataset$tempo != 0], dataset$popularity[dataset$tempo != 0],
     xlab = "Tempo (BPM)",
     ylab = "Popularity",
     main = "Popularity vs Tempo")

abline(
  lm(popularity ~ tempo, data = dataset),
  col = "red"
)

```

The correlation coefficient between popularity and tempo is 0.014. This is indicates a very small positive correlation, As tempo increases, popularity increases very slightly. This makes sense logically, as people often enjoy upbeat and up-tempo songs that make them feel good.

##### Correlation between Popularity and Mode

```{r}

cor(dataset$popularity, dataset$mode, use = "complete.obs")

dataset$mode_label <- factor(
  dataset$mode,
  levels = c(0, 1),
  labels = c("Minor", "Major")
)

boxplot(
  popularity ~ mode_label,
  data = dataset,
  col = c("lightpink", "lavender"),
  main = "Popularity by Musical Mode",
  xlab = "Mode",
  ylab = "Popularity"
)

```

The correlation coefficient is -0.013, indicating that there is a very slight negative correlation between mode and popularity. Songs in a major key are slightly less popular overall than songs in a minor key. This is surprising, since the major mode sounds happier, and you'd think people would rather listen to a song with a happier sound. However, making a song in a minor key is more difficult, so it's possible that they tend to have a better production due to being more challenging for less experienced musicians.

##### Correlation between Popularity and Loudness

```{r}

cor(dataset$popularity, dataset$loudness, use = "complete.obs")

plot(
  dataset$loudness,
  dataset$popularity,
  main = "Popularity vs Loudness",
  xlab = "Loudness",
  ylab = "Popularity",
  pch = 19,
)

abline(
  lm(popularity ~ loudness, data = dataset),
  col = "red",
  lwd = 2
)

```

The correlation coefficient is 0.05, which is the highest of all the correlations so far. While still a slight correlation, songs tend to be more popular as their loudness increases. According to this article (<https://pmc.ncbi.nlm.nih.gov/articles/PMC5634808/>), music arouses our brain, and loudness intensifies your response to the music. So if you like a song, you're more likely to want it to be louder to add to your experience.

##### Conclusion

Overall, most of the descriptive variables seemed to have very little effect on the popularity of the song. While there were some slight trends, popularity is likely far more dependant on the popularity of the artist and the perceived quality of the song, both things that are not tracked within this dataset.

#### What is the average of all of the other descriptive variables for each genre?

```{r}

library(dplyr)
library(knitr)

genres <- c(
  "acoustic",
  "country",
  "rock",
  "jazz",
  "metal",
  "edm",
  "pop",
  "hip-hop",
  "dance",
  "classical"
)

get_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

genre_summary <- dataset %>%
  filter(track_genre %in% genres, key != -1) %>%
  group_by(track_genre) %>%
  summarise(
    Tempo = round(mean(tempo, na.rm = TRUE), 1),

    Loudness = round(mean(loudness, na.rm = TRUE), 2),

    Popularity = round(mean(popularity, na.rm = TRUE), 1),

    `Explicit (%)` = round(mean(explicit, na.rm = TRUE) * 100, 1),

    `Duration (min)` = round(mean(duration_ms, na.rm = TRUE) / 60000, 2),

    Mode = ifelse(
      get_mode(mode) == 1,
      "Major",
      "Minor"
    ),

    `Time Sig.` = paste0(get_mode(time_signature), "/4"),

    Key = recode(
      as.character(get_mode(key)),
      "0" = "C",
      "1" = "C#/Db",
      "2" = "D",
      "3" = "D#/Eb",
      "4" = "E",
      "5" = "F",
      "6" = "F#/Gb",
      "7" = "G",
      "8" = "G#/Ab",
      "9" = "A",
      "10" = "A#/Bb",
      "11" = "B"
    ),

    .groups = "drop"
  )

knitr::kable(
  genre_summary,
  caption = "Summary Statistics by Genre",
  align = "c",
  format = "markdown"
)
```

In order to calculate the average descriptive variables of each genre, we have to assemble a list of diverse genres that somewhat represent the full list of 125. Then, we can create a table to view the averages for each of the descriptive variables.

- Tempo: The genre with the highest tempo on average is rock at 126.3, and the lowest is classical at 107.9.
- Loudness: The loudest genre on average is EDM with -5.28 decibels, and the quietest by far is classical at -20.15
- Popularity: The most popular genre is (predictably) pop, with an average score of 47.6. The least popular is classical, with an average of 13.1.
- Explicit (%): The genre with the highest percentage of explicit content is hip-hop at 31.9%, and the lowest percentage is classical at 0%. Now I'm not saying that someone should make a classical song with swear words, but I am saying that there's an untapped market there.
- Duration (min): The genre with the longest average duration in minutes is metal at 4.27, and the shortest genre is jazz at 3.17.
- Mode: The most common mode across all genres is Major.
- Time Signature: The most common time signature across all genres listed here is 4/4.
- Key: C#/Db is the most common key for dance, hip-hop, and jazz music. C is the most common key for acoustic and rock music. G is the most common key for classical and country music. B is the most common key for EDM, D is the most common for metal, and F is the most common for pop.

####  How do mood-related Spotify audio features differ across selected music genres?

For this research question, I focused on four Spotify audio features that are strongly related to the mood or feeling of a song: danceability, energy, tempo, and valence. Danceability measures how suitable a song is for dancing, energy measures how intense or active the song feels, tempo measures the speed of the song in beats per minute, and valence measures how positive or cheerful the song sounds. Instead of building a prediction model, this part of the project uses exploratory data analysis to compare how these mood-related features differ across selected genres.

I selected 10 genres from the full Spotify dataset: pop, rock, hip-hop, jazz, classical, country, metal, acoustic, dance, and EDM. These genres were chosen because they represent a wide variety of musical styles, from softer genres such as acoustic and classical to more energetic genres such as metal, EDM, and dance. The bar chart showing the number of songs by genre confirms that the selected dataset is fairly balanced. Most genres have close to 1,000 songs, while classical and dance have slightly fewer. Because the sample sizes are similar, the comparison across genres is more reliable and easier to interpret.

The boxplots comparing mood-related audio features show clear differences between genres. For danceability, hip-hop, dance, EDM, and pop tend to have higher median values, meaning songs in these genres are generally more suitable for dancing. This makes sense because these genres often use strong beats, repeated rhythms, and structures designed for movement. On the other hand, classical music has the lowest danceability values, which is expected because classical songs are usually not created for dancing in the same way modern popular genres are.

For energy, metal, EDM, dance, and rock generally have higher values. These genres often contain louder sounds, faster rhythms, stronger beats, or more intense instrumentation. Metal especially stands out as one of the most energetic genres. Classical, jazz, and acoustic songs tend to have lower energy values, showing that they are usually softer, calmer, or less intense. However, there is still overlap between genres, meaning not every song in a genre follows the same pattern.

For tempo, the differences between genres are not as strong as danceability or energy. Most genres have songs spread across a wide range of tempos, and many of the boxplots overlap. This suggests that tempo alone does not clearly separate genres. Even genres that feel different emotionally can have similar song speeds. However, EDM, dance, rock, and pop still show many songs with moderate to fast tempos, while classical and jazz include more songs with slower tempos.

For valence, which represents musical positivity, genres also show noticeable variation. Pop, dance, country, and rock tend to have more songs with medium to high valence, meaning they often sound more upbeat or positive. Metal has high energy but lower valence compared to some other genres, suggesting that it may sound intense but not always cheerful. Classical, acoustic, and jazz include many songs with lower or moderate valence, which may reflect more calm, emotional, or complex moods.

The scatterplot of energy versus valence gives another way to understand mood differences. Songs near the top of the plot have high energy, while songs farther to the right have higher positivity. The plot shows that many genres overlap, so energy and valence alone cannot perfectly separate genres. However, some general patterns can still be seen. Metal songs appear more often in the high-energy area, while classical songs are more common in the low-energy area. Pop, dance, rock, and country are more spread out, showing that they include both positive and less positive songs with different energy levels.

Finally, the stacked bar chart divides songs into four simple mood categories based on whether they are above or below the median for energy and valence: High Energy / Positive, High Energy / Less Positive, Low Energy / Positive, and Low Energy / Less Positive. This chart makes the mood profile of each genre easier to compare. Metal has a large proportion of high-energy songs, especially songs that are less positive. EDM also has many high-energy songs, which matches its strong electronic and dance-based style. Classical and acoustic have larger proportions of low-energy songs, especially low-energy and less-positive songs. Genres such as rock, hip-hop, dance, pop, and country have a more balanced mix, but many of them still show a strong share of high-energy or positive songs.

Overall, the charts show that mood-related Spotify features differ across genres in meaningful ways. Danceability and energy show the clearest differences, especially between modern rhythmic genres such as hip-hop, dance, EDM, and metal compared to softer genres such as classical and acoustic. Tempo is less useful for separating genres because many genres have overlapping tempo ranges. Valence helps show whether genres tend to sound more positive or less positive, but it also overlaps across genres. Therefore, mood in music is not determined by only one feature. Instead, the emotional feeling of a genre is better understood by looking at multiple audio features together.